%============== NÃO MODIFICAR ESTES COMANDOS =============
% Opções:
%   Qualificação          = qualificacao
%   Curso                 = doutorado/mestrado
%   Situação do trabalho  = pre-defesa/pos-defesa (exceto para qualificação)
%   Versão para impressão = impressao
%=========================================================

% QUALIFICAÇÃO DE DOUTORADO:
\documentclass[qualificacao,doutorado]{packages/icmc}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{url}
\usepackage{mdframed}
\usepackage{enumerate}
\usepackage[document]{ragged2e}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{algpseudocode}

% ------------------------------------------------------------------------
% Cover and title-page metadata
% ------------------------------------------------------------------------

\tituloPT{Minimapas de Código-Fonte como Abordagem Visual para Análise de Software}
\tituloEN{Source Code Minimaps as a Visual Approach for AI Analysis}

\autor[Gebara Junior, Munif]{Munif Gebara Junior}
\genero{M}

\orientador[Costa, Yandre Maldonato e Gomes da]{Prof. Dr.}{Yandre Maldonato e Gomes da Costa}
\iesorientador{Universidade Estadual de Maringá (DIN--UEM)}
\generoOR{M}

\coorientador[Wiese, Igor Scaliante and Cordeiro, André F. R.]{Prof. Drs.}{Igor Scaliante Wiese e André F. R. Cordeiro}
\iescoorientador{Universidade Tecnológica Federal do Paraná (UTFPR), Campus Campo Mourão, Brasil; e Universidade Estadual de Maringá (DIN--UEM)}
\generoCOOR{M}

\curso{CCMC}
\data{01}{07}{2026}
\idioma{EN}

% Qualification documents — approval page fields left neutral
\dataaprovacao{July 1, 2026}
\localdefesa{Universidade Estadual de Maringá}

\bancauemnomeA{}
\bancaueminstituicaoA{}
\generoUEMA{M}

\bancauemnomeB{}
\bancaueminstituicaoB{}
\generoUEMB{M}

\bancaexternonomeA{}
\bancaexternoinstituicaoA{}
\generoExteriorA{M}

\bancaexternonomeB{}
\bancaexternoinstituicaoB{}
\generoExteriorB{M}

% ------------------------------------------------------------------------
% Abstracts
% ------------------------------------------------------------------------

\textoresumo[english]{
\textbf{Context:} Modern software repositories are large, heterogeneous, and difficult to analyze at scale. A key barrier for industrial adoption of external analysis tools is that source code often encodes strategic business information in identifiers, numeric constants, formulas, and comments---content that organizations cannot share with external services. Many existing machine learning approaches retain this lexical layer, limiting their applicability in proprietary contexts. \textbf{Objective:} This work investigates source code minimaps as compact, anonymized visual signatures for software analysis. A minimap is generated by mapping each character in a source file to a grayscale pixel: control characters and whitespace map to intensity 0, digits to 53, uppercase letters to 77, lowercase letters to 109, and structural symbols to their raw ASCII value. This suppresses all identifiers and numeric literals---the lexical layer that carries intellectual property---while preserving structural signals such as indentation, punctuation density, operator placement, and block organization. The goal is to evaluate whether this compact visual abstraction supports scalable, language-agnostic software analysis and enables analysis of codebases that cannot be shared in raw form. \textbf{Methodology:} The research has been structured into three stages. The first study applied Local Binary Patterns and traditional classifiers (K-Nearest Neighbors, Random Forest, Support Vector Machine) to minimap images in a controlled setting with seven file-type classes. The second study scaled the approach to 685,906 minimap images extracted from 100 open-source projects across 10 programming languages, evaluating ResNet-18 and EfficientNet-B0 for three simultaneous classification objectives (Type, Project, Author) and using Integrated Gradients to interpret model decisions. The third stage produced MinimapSignatures, a tool ecosystem comprising a FastAPI/PyTorch inference backend, a Visual Studio Code extension, and an IntelliJ IDEA plugin that generate minimaps locally and transmit only the anonymized image---never the raw source---to the inference endpoint. \textbf{Preliminary results:} Random Forest over Local Binary Pattern features reached approximately 93\% F1-score on the initial controlled dataset. EfficientNet-B0 with ImageNet pretraining reached validation accuracies of 93.3\% (Type), 86.0\% (Project), and 87.1\% (Author) on the large-scale corpus. A preliminary comparison between anonymized and non-anonymized representations showed that predictive performance is largely preserved after anonymization, confirming that the discriminative signal resides in structural features rather than in lexical content. The Author objective, in particular, demonstrates that authorship style survives full lexical suppression---providing direct evidence that the model operates on structure, not on identifiers or literals. MinimapSignatures was functionally validated end-to-end; the quality classification target is reserved as a placeholder for future work. \textbf{Expected contributions:} The final thesis is expected to contribute a reproducible methodology for IP-preserving source code analysis, a formal anonymization ablation study with performance-vs-suppression trade-off curves, refined evaluation under project-disjoint dataset splits, a systematic comparison of handcrafted and learned visual representations, explainability protocols for visual software analysis models, and an extended MinimapSignatures prototype with multi-file support and developer-centered evaluation.
}{Source code minimaps; Software visualization; Machine learning; Intellectual property protection; Explainable artificial intelligence; IDE plugins}

% ------------------------------------------------------------------------
% Pre-textual elements
% ------------------------------------------------------------------------

\textodedicatoria*{tex/pre-textual/dedicatoria}
\textoagradecimentos*{tex/pre-textual/agradecimentos}
\textoepigrafe*{tex/pre-textual/epigrafe}

\incluilistadefiguras
\incluilistadetabelas
\incluilistadealgoritmos

\begin{document}

\textual

\chapter{Introduction}
\label{chapter:introduction}
\input{capitulos/1-Introducao.tex}

\chapter{Theoretical Foundations}
\label{chapter:foundations}
\input{capitulos/2-fundamentacao.tex}

\chapter{Related Work and Literature Mapping}
\label{chapter:related-work}
\input{capitulos/3-related_work.tex}

\chapter{Source Code Minimaps: Proposed Methodology}
\label{chapter:methodology}
\input{capitulos/4-methodology.tex}

\chapter{Preliminary Studies and Results}
\label{chapter:preliminary-results}
\input{capitulos/5-preliminary_results.tex}

\chapter{MinimapSignatures: Tool Support}
\label{chapter:tool-support}
\input{capitulos/6-tool_support.tex}

\chapter{Research Plan, Risks, and Threats to Validity}
\label{chapter:research-plan}
\input{capitulos/7-research_plan.tex}

\chapter{Final Remarks}
\label{chapter:final-remarks}
\input{capitulos/8-final_remarks.tex}

\bookmarksetup{startatroot}

\postextual

\bibliography{referencias}

\end{document}
