\section{Current Status}

At the qualification stage, the doctoral project has already produced three
main outputs. The first is an initial empirical study on source code minimaps,
Local Binary Patterns, and traditional classifiers \cite{gebara2025iwssip}. The
second is a large-scale study with 685,906 minimaps from 100 open-source
projects and 10 programming languages \cite{gebara2025ciarp}. The third is the
MinimapSignatures tool ecosystem, which validates the feasibility of bringing
minimap analysis into IDE workflows \cite{gebara2026minimapsignatures}.

These outputs establish a solid foundation, but the thesis is not complete. The
remaining work must deepen the methodology, broaden the task set, strengthen
evaluation, and connect the tool to developer-facing scenarios. This chapter
defines the planned research activities and the main risks.

\section{Expected Deliverables}

The final doctoral work is expected to produce a set of deliverables rather than
a single isolated experiment. The main deliverables are:

\begin{enumerate}
    \item a finalized methodology for source code minimap generation,
    anonymization, cataloging, and evaluation;
    \item a curated and documented minimap dataset or dataset-generation package;
    \item benchmark results comparing handcrafted descriptors and deep visual
    models under controlled splits;
    \item an anonymization study measuring the trade-off between information
    suppression and predictive performance;
    \item an explainability study showing which visual regions support model
    decisions;
    \item an extended MinimapSignatures prototype with multi-file or
    explanation-oriented feedback;
    \item at least one evaluation of developer-facing use, even if exploratory;
    \item the final thesis text and associated reproducibility package.
\end{enumerate}

These deliverables are designed to make the thesis coherent. The dataset and
methodology support the models. The models support the explanations. The
explanations support tool feedback. The tool feedback motivates developer
evaluation. Each part therefore contributes to the same research arc.

\section{Remaining Research Activities}

The remaining doctoral work is organized into six activities.

\subsection{A1: Refined Dataset Curation}

The large-scale dataset must be refined for additional experiments. This
includes improving label hygiene, identifying generated files, separating bot and
human contributors when possible, recording repository metadata, and creating
balanced or task-specific subsets. The catalog should support reproducible
filtering decisions.

This activity is necessary because the current dataset is large but imbalanced.
Head classes dominate some objectives, and Author labels may contain bots,
organizations, or aliases. Better curation will make later experiments more
trustworthy.

\subsection{A2: Anonymization Evaluation}

A preliminary comparison between anonymized and non-anonymized minimap
representations was conducted informally during the development of the approach.
Results showed minimal performance degradation, confirming that structural
signals---not lexical content---carry most of the discriminative information.

However, this comparison was not formally documented. Activity A2 will produce
a rigorous anonymization study by comparing several suppression levels:

\begin{itemize}
    \item baseline (current character-to-pixel mapping: letters→canonical,
    digits→canonical, others→raw ASCII);
    \item no anonymization (raw character codes, including identifiers and literals);
    \item comment removal (strip comment regions before rendering);
    \item punctuation normalization (map all operators and delimiters to a single
    canonical value, testing whether structure beyond punctuation still carries signal);
    \item maximum suppression (all non-whitespace characters to a single value,
    preserving only blank lines and indentation).
\end{itemize}

Each level will be evaluated on the same splits used in Study 2, reporting
accuracy, macro-F1, and per-class behavior. The goal is to produce a
performance-vs-suppression trade-off curve and to characterize which suppression
level is appropriate for contexts where intellectual property protection is the
primary concern.

\subsection{A3: Imbalance-Aware Modeling}

The next modeling phase will evaluate techniques designed for imbalanced
multi-class datasets. Candidate strategies include class weighting, balanced
sampling, focal loss, logit adjustment, mixup or cutmix variants, and deferred
reweighting schedules. Evaluation will prioritize macro-F1, per-class recall,
calibration, and confusion analysis, not only accuracy.

This activity responds directly to limitations observed in the large-scale
study. The goal is to improve minority-class behavior and to avoid drawing
conclusions from head-class performance alone.

\subsection{A4: Multi-Scale and Region-Aware Minimap Analysis}

Fixed 128 by 128 minimaps are compact and convenient, but they may discard
information from long files. Future experiments should compare multiple
resolutions, patch-based inputs, pyramidal representations, and region-aware
models. Explainability results can guide this process by showing which regions
are repeatedly salient.

This activity will help answer whether minimap performance is limited by input
resolution and whether region-level analysis can support more detailed tasks,
such as identifying suspicious sections of a file rather than classifying the
whole file.

\subsection{A5: Quality and Architectural-Deviation Tasks}

The thesis should move beyond Type, Project, and Author objectives toward tasks
closer to software quality. Candidate targets include code-smell likelihood,
test-file recognition, architectural role classification, generated-file
anomaly detection, and deviation from project-specific conventions.

This activity may combine minimap signals with existing static-analysis metrics
or labels. The goal is not to claim that minimaps alone can prove quality
problems, but to evaluate whether they can provide useful weak signals for
prioritization and review.

\subsection{A6: Tool Evaluation}

MinimapSignatures has been functionally validated, but it still requires
developer-centered evaluation. Future studies may involve controlled tasks,
think-aloud sessions, interviews, or scenario-based inspections. The evaluation
should measure usefulness, understandability, trust, and perceived fit with
developer workflows.

The tool should also be extended to multi-file analysis and explanation
overlays. These features are likely to make the tool more useful than single
file-level prediction alone.

\section{Planned Experiments}

Table~\ref{tab:planned-experiments} summarizes planned experiments and their
expected outcomes.

\begin{table}[ht]
\centering
\caption{Planned experiments for the doctoral thesis.}
\label{tab:planned-experiments}
\begin{tabular}{p{0.22\textwidth}p{0.35\textwidth}p{0.28\textwidth}}
\toprule
\textbf{Experiment} & \textbf{Description} & \textbf{Expected output} \\
\midrule
Anonymization ablation & Compare several anonymization modes and comment-removal strategies. & Trade-off curves between performance and information suppression. \\
Imbalance-aware learning & Evaluate class weights, balanced sampling, focal loss, and calibration. & Improved macro-F1 and minority-class analysis. \\
Resolution analysis & Compare 128 by 128, larger fixed sizes, and patch-based inputs. & Evidence about information loss and scalability. \\
Model comparison & Evaluate additional CNN backbones, transformer-based variants, and the multitask architecture under the same splits as Study 2. & Stronger benchmark of architecture choices. \\
Quality target pilot & Use labeled or proxy quality indicators for minimap-based classification. & Evidence about feasibility for quality-related analysis. \\
Tool study & Evaluate MinimapSignatures with developers or realistic inspection tasks. & Feedback on usefulness, trust, and workflow integration. \\
\bottomrule
\end{tabular}
\fautor
\end{table}

\section{Evaluation Plan}

The evaluation plan combines quantitative and qualitative evidence. Quantitative
model evaluation will use train/validation/test splits, multiple random seeds
when feasible, macro-F1, weighted-F1, accuracy, ROC-AUC, calibration measures,
and confusion matrices. When objectives involve repositories or authors, splits
must be designed to avoid leakage. For example, project-specific tasks must not
accidentally train and test on near-duplicate artifacts.

Qualitative evaluation will be used in two places. First, explainability maps
will be inspected to verify whether visual evidence is plausible and stable.
Second, tool evaluation will collect developer perceptions through interviews,
questionnaires, or observational protocols. These qualitative data will not
replace quantitative metrics; they will help interpret whether the tool's
feedback is understandable and useful.

Empirical software-engineering guidelines will inform reporting, including
explicit threats to validity and transparent description of datasets, procedures,
and limitations \cite{kitchenham2002preliminary,wohlin2012experimentation,runeson2012case}.

\section{Threats to Validity}

\subsection{Construct Validity}

Construct validity concerns whether the study measures what it claims to
measure. In this project, a major construct threat is the interpretation of
labels. Type labels may mix file extension with semantic role. Project labels
may capture boilerplate rather than meaningful project identity. Author labels
may reflect bots, generated files, or organizational accounts rather than human
style. Quality labels may be noisy if derived from static-analysis tools or
proxies.

Mitigation strategies include careful label documentation, sensitivity analysis,
manual inspection of samples, exclusion of ambiguous labels when needed, and
explicit discussion of what each objective actually represents.

\subsection{Internal Validity}

Internal validity concerns whether observed effects are caused by the method
rather than confounding factors. In minimap experiments, potential confounders
include duplicate files, generated artifacts, repository templates, file naming
patterns, and data leakage across splits. If training and test sets contain
near-identical generated files, performance may be inflated.

Mitigation strategies include content hashing, duplicate removal, repository-aware
splits, time-aware splits when possible, and ablation studies that remove
headers, comments, or generated files.

\subsection{External Validity}

External validity concerns whether results generalize beyond the evaluated
corpus. The large-scale dataset includes 100 open-source projects and 10
languages, but it may not represent proprietary systems, safety-critical
software, small scripts, educational code, or low-resource languages. Industrial
code may also have conventions not present in public repositories.

Mitigation strategies include expanding the corpus, reporting repository
characteristics, evaluating additional ecosystems, and conducting industrial or
semi-industrial case studies when access is possible.

\subsection{Conclusion Validity}

Conclusion validity concerns whether the statistical and analytical conclusions
are justified. Class imbalance, multiple comparisons, and single-run results can
lead to overconfident claims. Accuracy alone may hide poor minority-class
performance.

Mitigation strategies include reporting multiple metrics, confidence intervals
or variation across seeds when feasible, per-class results, macro-averaged
metrics, and clear distinction between preliminary and final evidence.

\section{Intellectual Property and Responsible Use}

The primary ethical motivation of this project is the protection of intellectual
property. Industrial codebases often contain strategic business information
embedded in identifiers, numeric constants, formulas, and comments. The
minimap approach is designed to enable structural analysis of source code without
exposing this information to external services or collaborators.

The character-level anonymization adopted here guarantees that identifier names,
numeric literals, and string content are suppressed before any minimap is
transmitted or stored externally. A preliminary experiment confirmed that this
suppression has minimal impact on predictive performance, which means the utility
of the analysis is preserved while the business-sensitive lexical layer remains
local.

The Author prediction objective is a methodological probe, not a surveillance
tool. Its purpose is to demonstrate that stylistic patterns---indentation habits,
spacing conventions, operator placement---survive anonymization. This is evidence
that the model operates on structure rather than on lexical content. Author
prediction is not proposed as a deployment target in personnel monitoring,
performance evaluation, or attribution of responsibility. Any use of this
objective outside its methodological role would be a misapplication of the
research.

For practical tool deployment, the following design principles apply:

\begin{enumerate}
    \item generate minimaps locally and transmit only the anonymized image, never
    the raw source file;
    \item make the anonymization transformation auditable and documented;
    \item allow individual objectives to be disabled at the user's discretion;
    \item communicate prediction results as probabilistic signals, not as facts;
    \item keep humans responsible for all decisions based on model output.
\end{enumerate}

The MinimapSignatures backend already implements local anonymization in the IDE
plugins: the minimap is generated client-side and only the image is sent to the
inference endpoint, not the source text. This design reflects the principle that
intellectual property protection begins at the data-collection step.

\section{Risks and Mitigation}

Table~\ref{tab:risks} summarizes the main research risks and mitigation actions.

\begin{table}[ht]
\centering
\caption{Research risks and mitigation strategies.}
\label{tab:risks}
\begin{tabular}{p{0.26\textwidth}p{0.33\textwidth}p{0.27\textwidth}}
\toprule
\textbf{Risk} & \textbf{Impact} & \textbf{Mitigation} \\
\midrule
Dataset imbalance dominates results & Inflated accuracy and weak minority-class performance. & Use macro metrics, reweighting, balanced subsets, and per-class analysis. \\
Anonymization removes too much signal & Models lose predictive power. & Preliminary comparison shows minimal degradation; formal ablation will quantify trade-off curves. \\
Stronger suppression needed for certain contexts & Current mapping may not satisfy all IP protection requirements. & Evaluate comment removal, punctuation normalization, and maximum suppression modes in Activity A2. \\
Models learn boilerplate artifacts & Poor generalization across repositories. & Use XAI, ablations, and repository-disjoint evaluation. \\
Tool feedback is not useful to developers & Weak practical contribution. & Conduct user-centered evaluation and refine interaction design. \\
Quality labels are noisy & Invalid quality-related conclusions. & Use multiple label sources and manual validation samples. \\
\bottomrule
\end{tabular}
\fautor
\end{table}

\section{Tentative Timeline}

Table~\ref{tab:timeline} presents a tentative schedule after the qualification
deposit. The schedule should be refined with the advisor and co-advisors after
the qualification exam.

\begin{table}[ht]
\centering
\caption{Tentative doctoral research timeline after qualification.}
\label{tab:timeline}
\begin{tabular}{p{0.22\textwidth}p{0.62\textwidth}}
\toprule
\textbf{Period} & \textbf{Main activities} \\
\midrule
July--August 2026 & Incorporate qualification feedback; freeze revised research questions and thesis structure. \\
September--December 2026 & Refine dataset catalog; run anonymization and imbalance-aware experiments. \\
January--April 2027 & Evaluate additional architectures, resolutions, and explanation protocols. \\
May--August 2027 & Design and execute quality or architectural-deviation pilot studies. \\
September--December 2027 & Extend MinimapSignatures for multi-file analysis and explanation overlays; conduct developer-centered evaluation. \\
January--March 2028 & Consolidate results, write thesis chapters, and prepare artifact documentation. \\
April--June 2028 & Revise thesis, submit papers/artifacts, and prepare final defense material. \\
\bottomrule
\end{tabular}
\fautor
\end{table}

The schedule is intentionally conservative. It assumes that the core
representation and preliminary studies are already available, but that the final
thesis still requires stronger evaluation, broader tasks, and tool-centered
evidence.
