\section{Summary}

This qualification document presented a doctoral research proposal centered on
source code minimaps as visual signatures for software analysis. The proposal
starts from a practical problem: modern software repositories are heterogeneous,
large, and difficult to understand at scale, while many analysis approaches
depend on language-specific processing, expert-defined rules, or representations
that preserve sensitive lexical content.

The proposed approach transforms source files into compact, anonymized minimap
images that preserve structural, spatial, and syntactic cues. These images can
be analyzed with handcrafted texture descriptors, deep learning models, and
explainable AI methods. The approach is not intended to replace traditional
static analysis, parsers, or software visualization. Rather, it offers a
complementary representation that can support classification, prioritization,
interpretation, and developer-facing feedback.

\section{Current Evidence}

The research has already produced preliminary evidence. The first empirical
study showed that Local Binary Patterns and traditional classifiers can detect
software features from source code minimaps, with Random Forest reaching
approximately 93\% performance in the evaluated setting. The second study scaled
the representation to 685,906 minimaps from 100 open-source projects across 10
programming languages and evaluated deep models for Type, Project, and Author
objectives. The tool-support line, MinimapSignatures, demonstrated that minimap
generation and inference can be integrated into Visual Studio Code and IntelliJ
IDEA through a shared backend.

These results support the central hypothesis that anonymized minimaps preserve
useful structural information for software analysis. They also reveal the main
remaining challenges: class imbalance, label quality, anonymization strength,
explainability, generalization, and developer-centered evaluation.

\section{Expected Thesis Contribution}

The expected thesis contribution is a coherent methodology and artifact set for
visual source code analysis with minimaps. This includes a representation, an
anonymization strategy, dataset-generation procedures, empirical evaluation,
explainability protocols, and IDE-integrated tooling. Scientifically, the thesis
aims to position source code minimaps as a reproducible bridge between software
visualization and machine learning for code. Practically, it aims to provide
tools and datasets that can be reused for future research on software
comprehension, quality assessment, architectural deviation, and code-review
support.

\section{Closing Statement}

The qualification stage shows that the project has moved beyond an initial idea:
it already includes empirical studies, a large dataset, explainability results,
and working tool prototypes. The remaining doctoral work will refine the
evidence, expand the task set, evaluate practical usefulness, and consolidate
the approach into a thesis. If successful, the project will contribute a
language-light and visually grounded way of analyzing software artifacts,
opening a path for source code analysis that is compact, reproducible,
interpretable, and closer to the environments where developers work.
